### 2024-04-24 papers/cs444/paper-cs444.tex
latex support in remote ssh

\documentclass[review,sigconf]{acmart}

\usepackage{graphicx}
% \graphicspath{{figures/}}
%\usepackage{multicol}
\usepackage{multirow}
\usepackage{balance}
\usepackage{amsmath}
\usepackage{framed}

\newcommand{\todo}[1]{\textcolor{red}{TODO: #1}}

\begin{document}

\title[Bayesian]{Bayesian Grading: Analysis of Scores in the Context of Cheating}
\author{Victor Zhao}
\email{chenyan4@illinois.edu}
%\orcid{1234-5678-9012}
\affiliation{%
	\institution{University of Illinois Urbana-Champaign}
	\country{}
}
\author{Yuxuan Chen}
\email{yuxuan19@illinois.edu}
%\orcid{1234-5678-9012}
\affiliation{%
	\institution{University of Illinois Urbana-Champaign}
	\country{}
}
\author{Evan Matthews}
\email{evanmm3@illinois.edu}
%\orcid{1234-5678-9012}
\affiliation{%
	\institution{University of Illinois Urbana-Champaign}
	\country{}
}

\begin{abstract}
\todo{Abstract}
\end{abstract}

\maketitle

\section{Introduction}
\todo{Introduction}

Something about automated assessments and testing, resource limitations, unproctoring~\cite{gulati2024regimens}, untrustable grades, and our problem.

\section{Background}
\subsection{Computer-Based Assessments}
% Autograded problems, automated item generation, everything enabling the use of computer-based testing. Asynchronous testing

In recent classroom settings, computer-based assessments has been observed to have many advantages over the traditional paper-based assessments~\cite{chen2019generation}. 
One property of computer-based assessments is the ability to autograde students' work and provide instant feedback. A large body of research has shown that providing feedback in time can be beneficial to student learning~\cite{anderson1995cognitive}. 
In formative contexts, the ability to autograde problems and provide instant feedback also enables a mastery approach in courses~\cite{west2015prairielearn}. 
With computer-based assessments, students also have the power to work on problems and submit their work asynchronously, enabling more flexibility in classrooms~\cite{zhao2023flexibility}. 
Automatic Item Generation~\cite{chen2019generation} allows computers to generate problems that are similar to each other based on item models, providing a large problem pool students can practice with. 
The combination of Automatic Item Generation and autograding also allows for more courses to be offered at scale through online open assessment platforms, as the number of course enrollments increase significantly in recent years~\cite{silva2020measuring}. 
New technologies in development such as Automatic Short Answer Grading~\cite{bonthu2021automated} are expanding the use for computer-based assessments to more types of problems. These features can enable computer-based assessments in almost any discipline.

\subsection{Computer-Based Testing}
% CBTF, frequent testing, lack of resources, some unproctoring, cheating

Computer-based assessments can be used in summative testing contexts as well. Dishonesty is usually a large factor of the effectiveness of summative assessments~\cite{arnold2016cheating}. 
Compared to the traditional pen and paper exams, students have more control over the environment in which they are taking the exam, but plagiarism and cheating prevention can also be more challenging because students can potentially have more resources (legal or illegal) available to them~\cite{kurniawan2020byod}. 
Some of the initial proctoring methods is to 
Proctoring the exams at a large scale poses a lot of logistical problems such as the trouble students need to go through to setup a testing environment using lockdown browsers or other similar technologies~\cite{kurniawan2020byod}, and the difficulty in conducting frequent testing for classes and dealing with accomodations or conflict exams~\cite{zilles2015cbtf}. 
Training the course staff to be responsible proctors could also be time-consuming.

One solution proposed by researchers is to have a dedicated computer laboratory specifically for computer-based testing. 
This method was adopted and used to build the Computer-Based Testing Facility (CBTF) at the University of Illinois Urbana-Champaign, which has been utilized for an extended period of time~\cite{zilles2015cbtf}. 
The CBTF was built in a laboratory using computers configured for testing environments and hiring highly-trained proctors. 
This method has shown to be reasonable in terms of cost~\cite{zilles2015cbtf}, and also effective in preventing cheating on a large scale~\cite{silva2020measuring}.

However, with more classes incorporating computer-based assessments and computer-based testing, there might not be enough resources to have all exams proctored through the CBTF~\cite{mcmahon2022lessons}.

\subsection{Bayesian?}

\section{Experiment}
\subsection{Context}
The data for this study was drawn from a lower-division undergraduate computer science (CS) course over two semesters: Fall 2021 and Fall 2023.
Students were expected to have completed an introductory algebra course as a prerequisite prior to taking this course.
The course included 4 unproctored quizzes, 3 proctored exams, and 1 comprehensive proctored final exam.
Each quiz and exam had a duration of 50 minutes, except for the longer, cumulative final exam.
The quizzes were given one week before the exams and had overlapping content, allowing exams to serve as a rough mirrored comparison to the quiz just before the exam.
Students took the proctored exams within the CBTF. % synchronously or asynchronously?
We gathered exam records for the 695 students enrolled in Fall 2021 and the 423 students enrolled in Fall 2023.
After excluding students with test accommodations or those who didn't complete all the exams or quizzes, the final dataset included 622 student records from Fall 2021 and 405 student records from Fall 2023.

\subsection{Method}
Our proposed method analyzes student scores through policy simulation of computed Bayesian probabilities.
First, probabilities of student scores are computed with respect to their observed ability \(\mu\in \{0, 1, ..., 100\}\) and cheating boolean
\(c \in \{0, 1\}\) where \((c = 0), (c = 1)\) indicate not cheating and cheating, respectively. These parameters are used to compute
\begin{equation}
P(\text{obs} \mid \mu, c) = \frac{1}{N}\exp\bigg(-\frac{1}{2}\sum_{i=1}^{n} \left( \frac{a_i - \mu - \chi c}{\sigma} \right)^{2} \bigg)
\end{equation}
where \(a_i\) is the \(i\)th assessment of a student with parameters \((\mu, c)\), \(\sigma\) is the standard deviation of a student's assessments, and \(\chi\) is a nonnegative "penalty" applied
when a student has cheated. In order to distribute our data across student ability and/or cheating booleans, we use Bayes' Theorem to invert the relationship
of our computed probabilities:
\begin{equation}
	P(\mu, c \mid \text{obs}) = \frac{P(\text{obs}\mid (\mu, c)) \cdot P(\mu, c)}{\sum_{\mu', c'}P(\text{obs} \mid \mu', c')}
\end{equation}
where \(P(\mu, c)\) is the probability of a student's performance amongst their class \footnote[1]{A uniform distribution of student performances can also be used here.}, 
and our probability is normalized by the sum of previously computed probabilities \(P(\text{obs}\mid \mu, c)\). The follow computations returns a
probability array of shape \((mu, c)\) from which we can visualize the distributions of cheating and non-cheating students.

finally, we define a policy \(\pi_\theta(p)\) for simulating random variable \(S = (\mu, c)\) and testing whether the correct outcome was returned:
\begin{equation}
	\pi_\theta(p) = 
	\begin{cases}
		0 & p \le \theta\\
		1 & p > \theta 
	\end{cases}
\end{equation}
The result of performing this simulation \(m\) times is a \(2\times 2\) truth table indicating percentages of true and false outcomes of cheating and non-cheating for the data.
\subsection{Analysis}

\section{Results}

\section{Discussion}
\subsection{Limitations}
While existing datasets of student assessments confirm the viability of Bayesian analysis, such information greatly reduces the extent to which
the statistics can be considered trustworthy. In particular, our aforementioned analysis relies on the data being summative and free of
statistically significant cheating attempts. Bayesian analysis on formative assessments is entirely possible, but the computation time drastically
increases with respect to the context for which the analysis is performed. A weekly topic-by-topic basis is likely to provide more results with finer
details given enough time, but two to three summative assessments can return similar results by individual problems while significantly decreasing
computation time. Additional cheating attempts can also reduce the initial trustworthiness of scores and diminish Bayesian analysis's effectiveness.
As such, our results appear highly volatile or inconclusive without a subset of reasonable scores for analysis.

\section{Conclusions}

\bibliographystyle{ACM-Reference-Format}
\bibliography{references}

\end{document}

### 2024-04-24 papers/cs444/paper-cs444.tex
start of cs444 research report

@@
 
-\title[Bayesian]{Bayesian Grading: Analysis of Scores in the Context of Cheating}
-\author{Victor Zhao}
-\email{chenyan4@illinois.edu}
-%\orcid{1234-5678-9012}
-\affiliation{%
-	\institution{University of Illinois Urbana-Champaign}
-	\country{}
-}
-\author{Yuxuan Chen}
-\email{yuxuan19@illinois.edu}
-%\orcid{1234-5678-9012}
-\affiliation{%
-	\institution{University of Illinois Urbana-Champaign}
-	\country{}
-}
+\title[omr-lilypond-midi]{Optical Music Recognition for MIDI-LilyPond File Generation}
 \author{Evan Matthews}
@@
 
-Something about automated assessments and testing, resource limitations, unproctoring~\cite{gulati2024regimens}, untrustable grades, and our problem.
+In the world of music or audio transcription/arrangement, the complexity of sound and score data has allowed human performance to remain the current state-of-the-art.
+Several factors contribute to this circumstance: the number of potential audio and score file types, ambiguity on how performance qualities are notated, and overall inconsistencies between recordings and their respective scores.
+Machine learning models, in turn, have been a crucial step towards reducing these inconsistencies.
+Their ability to learn the nonlinearities and artistic qualities that otherwise plague audio computation have allowed for noteworthy advancements such as the WaveNet\cite{oord2016wavenet}.
+The remaining issue in the process of sound generation is the amount of data available to train with.
+Worthwhile audio data remains difficult to collect due to its large size and potential copyright issues.
+In particular, trying to condition off another medium is incredibly difficult as current datasets lack the correlations necessary to streamline the audio transcription process.
+Datasets such as MAESTRO are close to ideal results, but some intermediate steps are needed to learn correlations between audio files and musical scores.
+
+With the following conditions in mind, I propose an Image-to-MIDI model to serve a few purposes.
+First, this model will serve to convert image data to relative musical data to a high degree of accuracy.
+The process of converting MIDI to an image is trivial, but the opposite is a known Optical Music Recognition (OMR) problem in that the correlation is nonexistent.
+Second, a highly reliable model of this type will be capable of serving as an intermediate step in future audio computation endeavors, as the nontrivial nature of OMR is a barrier for reliable model training between scores and other types of audio data.
+Finally, a conversion of this type allows for more flexible datasets because MIDI can be converted into a number of other audio file types.
+
+From a technical standpoint, my project will consist of: the construction of a trained model to relate images to MIDI data, a paper describing the process of constructing the model, and any notable results and accuracies, and a few demos from which the accuracy of the model can also be visually confirmed.
+The model will generally reference existing OMR research and will serve as an attempt to construct a completely new CNN model, (not be an implementation of an existing paper).
+Existing MIDI datasets have been narrowed down to strictly two-stave piano works by Johann Sebastian Bach to maintain rhythmic and compositional uniformity, although a generalized model for various instruments and number of staves can theoretically be produced given a large enough dataset.
+
+Along with serving as my four-credit-hour project for the course, this model will serve as a supplementary step in my MS thesis- a separately-proposed model for generating audio conditioned on sheet music, advised by Paris Smaragdis.
+I can confirm that neither project will trivialize the other, and the flexibility of this project will open the door for other interesting ML-audio research in the future.
+For further results, I may pursue auto-transcription with this model to further differentiate from my planned thesis.
 
 \section{Background}
-\subsection{Computer-Based Assessments}
-% Autograded problems, automated item generation, everything enabling the use of computer-based testing. Asynchronous testing
+\subsection{Optical Music Recognition}
+\todo{OMR background}
 
-In recent classroom settings, computer-based assessments has been observed to have many advantages over the traditional paper-based assessments~\cite{chen2019generation}. 
-One property of computer-based assessments is the ability to autograde students' work and provide instant feedback.
-A large body of research has shown that providing feedback in time can be beneficial to student learning~\cite{anderson1995cognitive}.
-In formative contexts, the ability to autograde problems and provide instant feedback also enables a mastery approach in courses~\cite{west2015prairielearn}. 
-With computer-based assessments, students also have the power to work on problems and submit their work asynchronously, enabling more flexibility in classrooms~\cite{zhao2023flexibility}. 
-Automatic Item Generation~\cite{chen2019generation} allows computers to generate problems that are similar to each other based on item models, providing a large problem pool students can practice with.
-The combination of Automatic Item Generation and autograding also allows for more courses to be offered at scale through online open assessment platforms, as the number of course enrollments increase significantly in recent years~\cite{silva2020measuring}.
-New technologies in development such as Automatic Short Answer Grading~\cite{bonthu2021automated} are expanding the use for computer-based assessments to more types of problems.
-These features can enable computer-based assessments in almost any discipline.
-
-\subsection{Computer-Based Testing}
-% CBTF, frequent testing, lack of resources, some unproctoring, cheating
-
-Computer-based assessments can be used in summative testing contexts as well. Dishonesty is usually a large factor of the effectiveness of summative assessments~\cite{arnold2016cheating}. 
-Compared to the traditional pen and paper exams, students have more control over the environment in which they are taking the exam, but plagiarism and cheating prevention can also be more challenging because students can potentially have more resources (legal or illegal) available to them~\cite{kurniawan2020byod}.
-Some of the initial proctoring methods is to 
-Proctoring the exams at a large scale poses a lot of logistical problems such as the trouble students need to go through to setup a testing environment using lockdown browsers or other similar technologies~\cite{kurniawan2020byod}, and the difficulty in conducting frequent testing for classes and dealing with accomodations or conflict exams~\cite{zilles2015cbtf}.
-Training the course staff to be responsible proctors could also be time-consuming.
-
-One solution proposed by researchers is to have a dedicated computer laboratory specifically for computer-based testing. 
-This method was adopted and used to build the Computer-Based Testing Facility (CBTF) at the University of Illinois Urbana-Champaign, which has been utilized for an extended period of time~\cite{zilles2015cbtf}.
-The CBTF was built in a laboratory using computers configured for testing environments and hiring highly-trained proctors. 
-This method has shown to be reasonable in terms of cost~\cite{zilles2015cbtf}, and also effective in preventing cheating on a large scale~\cite{silva2020measuring}.
-
-However, with more classes incorporating computer-based assessments and computer-based testing, there might not be enough resources to have all exams proctored through the CBTF~\cite{mcmahon2022lessons}.
-
-\subsection{Bayesian?}
+\subsection{Lilypond, MIDI}
 
 \section{Experiment}
-\subsection{Context}
-The data for this study was drawn from a lower-division undergraduate computer science (CS) course over two semesters: Fall 2021 and Fall 2023.
-Students were expected to have completed an introductory algebra course as a prerequisite prior to taking this course.
-The course included 4 unproctored quizzes, 3 proctored exams, and 1 comprehensive proctored final exam.
-Each quiz and exam had a duration of 50 minutes, except for the longer, cumulative final exam.
-The quizzes were given one week before the exams and had overlapping content, allowing exams to serve as a rough mirrored comparison to the quiz just before the exam.
-Students took the proctored exams within the CBTF. % synchronously or asynchronously?
-We gathered exam records for the 695 students enrolled in Fall 2021 and the 423 students enrolled in Fall 2023.
-After excluding students with test accommodations or those who didn't complete all the exams or quizzes, the final dataset included 622 student records from Fall 2021 and 405 student records from Fall 2023.
-
-\subsection{Method}
-Our proposed method analyzes student scores through policy simulation of computed Bayesian probabilities.
-First, probabilities of student scores are computed with respect to their observed ability \(\mu\in \{0, 1, ..., 100\}\) and cheating boolean
-\(c \in \{0, 1\}\) where \((c = 0), (c = 1)\) indicate not cheating and cheating, respectively. These parameters are used to compute
-\begin{equation}
-P(\text{obs} \mid \mu, c) = \frac{1}{N}\exp\bigg(-\frac{1}{2}\sum_{i=1}^{n} \left( \frac{a_i - \mu - \chi c}{\sigma} \right)^{2} \bigg)
-\end{equation}
-where \(a_i\) is the \(i\)th assessment of a student with parameters \((\mu, c)\), \(\sigma\) is the standard deviation of a student's assessments, and \(\chi\) is a nonnegative "penalty" applied
-when a student has cheated. In order to distribute our data across student ability and/or cheating booleans, we use Bayes' Theorem to invert the relationship
-of our computed probabilities:
-\begin{equation}
-	P(\mu, c \mid \text{obs}) = \frac{P(\text{obs}\mid (\mu, c)) \cdot P(\mu, c)}{\sum_{\mu', c'}P(\text{obs} \mid \mu', c')}
-\end{equation}
-where \(P(\mu, c)\) is the probability of a student's performance amongst their class \footnote[1]{A uniform distribution of student performances can also be used here.}, 
-and our probability is normalized by the sum of previously computed probabilities \(P(\text{obs}\mid \mu, c)\). The follow computations returns a
-probability array of shape \((mu, c)\) from which we can visualize the distributions of cheating and non-cheating students.
-
-finally, we define a policy \(\pi_\theta(p)\) for simulating random variable \(S = (\mu, c)\) and testing whether the correct outcome was returned:
-\begin{equation}
-	\pi_\theta(p) = 
-	\begin{cases}
-		0 & p \le \theta\\
-		1 & p > \theta 
-	\end{cases}
-\end{equation}
-The result of performing this simulation \(m\) times is a \(2\times 2\) truth table indicating percentages of true and false outcomes of cheating and non-cheating for the data.
-\subsection{Analysis}
+\subsection{Key resources}
+The Image-to-MIDI convolutional neural network will be built from scratch while referencing promising results from OMR and image-to-text research, (should the CNN need fine-tuning or other enhancement).
+Currently, the model dataset is the complete Bach Midi Index and will be narrowed down to fit additional constraints: two staves, single instrument, rhythmic quantization at the 16th or 32nd- note level.
+Model testing will focus primarily on self-trained work, but I am interested in comparing against pre-trained models if I have extra time.
+For hardware, training will take place on a Linux GPU cluster, courtesy of Paris Smaragdis and the UIUC-CS Audio Computing Lab.
+Programming work is using a combination of Python with Pytorch for the CNN, and Bash scripts with Lilypond to prepare and normalize the dataset.
 
@@
 \section{Discussion}
-\subsection{Limitations}
-While existing datasets of student assessments confirm the viability of Bayesian analysis, such information greatly reduces the extent to which
-the statistics can be considered trustworthy. In particular, our aforementioned analysis relies on the data being summative and free of
-statistically significant cheating attempts. Bayesian analysis on formative assessments is entirely possible, but the computation time drastically
-increases with respect to the context for which the analysis is performed. A weekly topic-by-topic basis is likely to provide more results with finer
-details given enough time, but two to three summative assessments can return similar results by individual problems while significantly decreasing
-computation time. Additional cheating attempts can also reduce the initial trustworthiness of scores and diminish Bayesian analysis's effectiveness.
-As such, our results appear highly volatile or inconclusive without a subset of reasonable scores for analysis.
+\todo{discussion}
 
 \section{Conclusions}
+\todo{conclusions}
 

### 2024-04-25 papers/cs444/paper-cs444.tex
citations added

@@
 \todo{Introduction}
+~\cite{oord2016wavenet,contreras2023omrcnn,contreras2023omrpiano,mayer2024practical,andrea2021note}
 
@@
 Machine learning models, in turn, have been a crucial step towards reducing these inconsistencies.
-Their ability to learn the nonlinearities and artistic qualities that otherwise plague audio computation have allowed for noteworthy advancements such as the WaveNet\cite{oord2016wavenet}.
+Their ability to learn the nonlinearities and artistic qualities that otherwise plague audio computation have allowed for noteworthy advancements such as the WaveNet~\cite{oord2016wavenet}.
 The remaining issue in the process of sound generation is the amount of data available to train with.
@@
 \bibliographystyle{ACM-Reference-Format}
-\bibliography{references}
+\bibliography{references-cs444}
 

### 2024-05-07 model.py
model framework

@@ -1,10 +1,34 @@
 import torch
+from torch import nn
+from torch.nn import functional as F
+from torchvision.models import resnet18, ResNet18_Weights
+from torch import Tensor
 
 
-class OMR_Network(torch.nn.Module):
+class CRNN_Network(torch.nn.Module):
     def __init__(self):
-        super(OMR_Network, self).__init__()
-        pass
+        super(CRNN_Network, self).__init__()
 
-    def forward():
-        pass
+        self.device = torch.device("cuda" if torch.cuda.is_available() else "cpu")
+
+        self.conv1 = nn.Conv2d()
+        self.pool1 = nn.MaxPool2d()
+        self.conv2 = nn.Conv2d()
+        self.pool2 = nn.MaxPool2d()
+        self.conv3 = nn.Conv2d()
+        self.conv4 = nn.Conv2d()
+        self.pool3 = nn.MaxPool2d()
+        self.conv5 = nn.Conv2d()
+        self.batch_norm1 = nn.BatchNorm2d()
+        self.conv6 = nn.Conv2d()
+        self.batch_norm2 = nn.BatchNorm2d()
+        self.pool4 = nn.MaxPool2d()
+        self.conv7 = nn.Conv2d()
+        # Map to sequence
+        self.lstm1 = nn.LSTM(bidirectional=True)
+        self.lstm2 = nn.LSTM(bidirectional=True)
+        # Transcription
+        self.dense = nn.Linear()
+
+    def forward(input):
+        input = 6

### 2024-05-07 papers/cs444/paper-cs444.tex
paper work

@@
 With the following conditions in mind, I propose an Image-to-MIDI model to serve a few purposes.
-First, this model will serve to convert image data to relative musical data to a high degree of accuracy.
+First, this model serves to convert image data to relative musical data to a high degree of accuracy.
 The process of converting MIDI to an image is trivial, but the opposite is a known Optical Music Recognition (OMR) problem in that the correlation is nonexistent.
-Second, a highly reliable model of this type will be capable of serving as an intermediate step in future audio computation endeavors, as the nontrivial nature of OMR is a barrier for reliable model training between scores and other types of audio data.
+Second, a highly reliable model of this type is capable of serving as an intermediate step in future audio computation endeavors, as the nontrivial nature of OMR is a barrier for reliable model training between scores and other types of audio data.
 Finally, a conversion of this type allows for more flexible datasets because MIDI can be converted into a number of other audio file types.
 
-From a technical standpoint, my project will consist of: the construction of a trained model to relate images to MIDI data, a paper describing the process of constructing the model, and any notable results and accuracies, and a few demos from which the accuracy of the model can also be visually confirmed.
-The model will generally reference existing OMR research and will serve as an attempt to construct a completely new CNN model, (not be an implementation of an existing paper).
+This model generally references existing OMR research and serves as an attempt to construct a completely new CNN model.
 Existing MIDI datasets have been narrowed down to strictly two-stave piano works by Johann Sebastian Bach to maintain rhythmic and compositional uniformity, although a generalized model for various instruments and number of staves can theoretically be produced given a large enough dataset.
-
-Along with serving as my four-credit-hour project for the course, this model will serve as a supplementary step in my MS thesis- a separately-proposed model for generating audio conditioned on sheet music, advised by Paris Smaragdis.
-I can confirm that neither project will trivialize the other, and the flexibility of this project will open the door for other interesting ML-audio research in the future.
-For further results, I may pursue auto-transcription with this model to further differentiate from my planned thesis.
 
@@
 
+\subsection{Dynamic quantization}
+\todo{quantization stuff}
+
+\subsection{CRNN Model}
+\todo{CRNN model}
+
 \section{Results}

### 2024-05-09 papers/cs444/paper-cs444.tex
paper progress, lilypond explanation

@@
 \usepackage{graphicx}
-% \graphicspath{{figures/}}
-%\usepackage{multicol}
+\usepackage{listings}
 \usepackage{multirow}
@@
 \usepackage{framed}
+\usepackage{subfigure}
 
@@
 
-\title[omr-lilypond-midi]{Optical Music Recognition for MIDI-LilyPond File Generation}
+\title[omr-lilypond-midi]{Optical Music Recognition for LilyPond File Generation}
 \author{Evan Matthews}
@@
 \todo{Introduction}
-~\cite{oord2016wavenet,contreras2023omrcnn,contreras2023omrpiano,mayer2024practical,andrea2021note}
+\cite{oord2016wavenet,contreras2023omrcnn,contreras2023omrpiano,mayer2024practical,andrea2021note}
 
@@
 
-\subsection{Lilypond, MIDI}
+\subsection{Lilypond}
+Lilypond was chosen for dataset file representation as it maintains accurate translation between digital (MIDI) and visual (image) notation systems~\cite{lilypond}.
+In particular, translation from Lilypond to MIDI is trivial, and Lilypond provides a more intuitive representation of musical information compared to raw bytes of MIDI data.
+This notation language consists of multiple layers to separate score components:
+
+\begin{itemize}
+		\item \textbf{Document level:} components related to page layout and high-level musical details, (number of instruments/tracks, score engraving information).
+		\item \textbf{Music level:} components related to low-level musical details, (notes, rhythms, durations, key/time signatures, tempo).
+\end{itemize}
+
+At the document level, \todo{document}
+
+\begin{figure*}
+	\begin{subfigure}
+		\centering
+		\includegraphics[width = .8\linewidth]{./figures/lilypond-snippet.png}
+		% \caption{Rendered Lilypond (.ily) file with code}
+		\label{fig:lilypond-snippet}
+	\end{subfigure}
+	~
+	\begin{subfigure}
+		\centering
+		\includegraphics[width = .8\linewidth]{./figures/lilypond-code.png}
+		\label{fig:lilypond-code}
+	\end{subfigure}
+	\caption{Lilypond (.ily) code and rendering, (time signature = 4/4)}
+	\label{fig:lilypond-example}
+\end{figure*}
+
+At the music level, rhythmic musical symbols are notated according to their pitches and relative time duration.
+pitches are all characters ${a, b, c, d, e, f, g, r}$, where $r$ is a "rest" meaning no pitch occurs.
+To ascend or descend in pitch beyond a single musical "octave," the $'$ and $,$ characters are appended to indicate one or more octave ascendings or descendings, respectively.
+Pitches are also proceeded by a rhythmic value representing its duration in time.
+These values are typically powers of two, (but can technically be any positive floating-point value), and they dictate how long a pitch is played with respect to the piece's time signature.
+For example, a score with time signature = 2/4 indicates that each measure has two beats, and each beat is the length of a quarter note (hard-coded in the denominator).
+In this case, the line $"c4 d4"$ would represent two quarter (4) notes or a single measure in the provided time signature.
+Additionally, notes without defined durational values take on the previous note's duration in a line, so a series of equal-duration notes is represented by one durational value.
+Finally, separations between measures are made with $"| \% n"$, marking the end of measure $n$.
+Figure~\ref{fig:lilypond-example} compares the rendering of a single staff of music with its relative Lilypond notation .
+
+% \begin{lstlisting}
+% trackBchannelB = \relative c {
+% 	r16 c' d e f d e c g'8 c b32 a b16 c8 
+% 	| % 2
+% 	d16 g, a b c a b g d'8 g f32 e f16 g8 
+% 	| % 3
+% }
+% \end{lstlisting}
 
@@
 \subsection{Key resources}
+\todo{Remove this section?}
 The Image-to-MIDI convolutional neural network will be built from scratch while referencing promising results from OMR and image-to-text research, (should the CNN need fine-tuning or other enhancement).

### 2024-05-09 quant_parse.py

@@ -34,6 +34,6 @@ def quant_parse(num: int = 1, path: str = './bach/invent/invent') -> int:
     return max_quant * numerator * int(denominator / 4)
 
 
-if __name__ == '__main__':
-    q = quant_parse(1)
-    print(q)
+# if __name__ == '__main__':
+#     q = quant_parse(1)
+#     print(q)

### 2024-05-09 papers/cs444/paper-cs444.tex
document-level

@@
 
-At the document level, \todo{document}
+At the document level, aspects of a score that stay mostly or completely constant throughout a piece are indicated by specific, indent-sensitive keywords.
+For example, the number of voices/instruments and respective number of staves are initialized with the $\backslash{Voice}$ and $\backslash{context}$ commands,
+while high-level musical information is initialized by commands such as $\backslash{time}$ for time signature, $\backslash{key}$ for key signature, and $\backslash{tempo}$ for performance speed in beats per minute.
 

### 2024-05-09 papers/cs444/paper-cs444.tex
big work

@@
 
-\title[omr-lilypond-midi]{Optical Music Recognition for LilyPond File Generation}
+\title[omr-lilypond-midi]{Optical Music Recognition for Homophonic LilyPond File Generation}
 \author{Evan Matthews}
@@
 \section{Introduction}
-\todo{Introduction}
-\cite{oord2016wavenet,contreras2023omrcnn,contreras2023omrpiano,mayer2024practical,andrea2021note}
+\subsection{Initial Proposal}
+\todo{Introduction citations}
+\cite{contreras2023omrcnn,,andrea2021note}
 
@@
 In particular, trying to condition off another medium is incredibly difficult as current datasets lack the correlations necessary to streamline the audio transcription process. 
-Datasets such as MAESTRO are close to ideal results, but some intermediate steps are needed to learn correlations between audio files and musical scores.
+Datasets such as MAESTRO ~\cite{hawthorne2018enabling} are close to ideal results, but some intermediate steps are needed to learn correlations between audio files and musical scores.
 
@@
 
-This model generally references existing OMR research and serves as an attempt to construct a completely new CNN model.
-Existing MIDI datasets have been narrowed down to strictly two-stave piano works by Johann Sebastian Bach to maintain rhythmic and compositional uniformity, although a generalized model for various instruments and number of staves can theoretically be produced given a large enough dataset.
+\subsection{Known Limitations}
+
+Given the vast complexity of western music notation and intricacies in notation/time alignment, the proposed model has a pessimistic approach. 
+That is, given a small subset of musical data (\textit{among all notated music}) and some model, results can be expected to only match expected outputs through one or more of: pitch, rhythm, duration, and overall formatting.
+Existing research in OMR supports these expectations. First, several digital notation systems exist for music, including:
+
+\begin{itemize}
+	\item Standards like \textit{MusicXML}, \textit{Lilypond} and \textit{MIDI},
+	\item Software-specific formats from \textit{Musescore}, \textit{Sibelius} and \textit{Finale},
+	\item \textit{KERN} from the Humdrum tool-set ~\cite{contreras2023omrpian},
+	\item Mayer, et al.'s \textit{Linearized MusicXML} ~\cite{mayer2024practical}, and
+	\item Contreras, et al.'s untitled "end-to-end OMR" encoding language ~\cite{contreras2023omrcnn}.
+\end{itemize}
+
+Each format provides benefits and drawbacks towards generalized OMR, but combined research efforts have yet to hone in on a particular format.
+Second, current research continues to limit its effective musical scope: constant genre, time period composed, single vs. multi-line pieces, and single vs. multiple measures, to name a few.
+These limitations are to be expected as a means of balancing experiment accuracy with the subset of musical notation to be recognized. 
+However, no paper to date has intended to capture a high accuracy while completely generalizing the space of recognizable music.
+Finally, the state-of-the-art for machine-learning-based OMR lies in the implementation of Convolutional Recurrent Neural Networks (CRNN). 
+This model type is preferred over CNN for its ability to learn data as order-dependent sequences- a crucial philosophy in parsing and understanding musical scores in general.
+
+\subsection{Research Questions}
+
+Despite low research expectations, I believe that a few questions can be posed and answered for the purpose of bounding requirements on a larger, all-encompassing Image-to-MIDI model:
+
+\begin{itemize}
+	\item [\textbf{RQ1}] What forms of recognition can be expected from a generalized CRNN implementation?
+	\item [\textbf{RQ2}] How does Lilypond recognition compare against other standard and custom formats?
+	\item [\textbf{RQ3}] What conflicts currently prevent state-of-the-art models from further generalization?
+\end{itemize}
 
@@
 
-\begin{figure*}
+\begin{figure}
 	\begin{subfigure}
@@
 	\end{subfigure}
-	\caption{Lilypond (.ily) code and rendering, (time signature = 4/4)}
+	\caption{Lilypond (.ly) code and rendering, (time signature = 4/4)}
 	\label{fig:lilypond-example}
-\end{figure*}
+\end{figure}
 
@@
 \section{Experiment}
-\subsection{Key resources}
-\todo{Remove this section?}
-The Image-to-MIDI convolutional neural network will be built from scratch while referencing promising results from OMR and image-to-text research, (should the CNN need fine-tuning or other enhancement).
-Currently, the model dataset is the complete Bach Midi Index and will be narrowed down to fit additional constraints: two staves, single instrument, rhythmic quantization at the 16th or 32nd- note level.
-Model testing will focus primarily on self-trained work, but I am interested in comparing against pre-trained models if I have extra time.
-For hardware, training will take place on a Linux GPU cluster, courtesy of Paris Smaragdis and the UIUC-CS Audio Computing Lab.
-Programming work is using a combination of Python with Pytorch for the CNN, and Bash scripts with Lilypond to prepare and normalize the dataset.
-
-\subsection{Dynamic quantization}
-\todo{quantization stuff}
 
@@
 \todo{CRNN model}
+The model in question
+
+\subsection{Data Specfications}
+\todo{dataset specifics}
+Existing MIDI datasets have been narrowed down to strictly two-stave piano works by Johann Sebastian Bach to maintain rhythmic and compositional uniformity, although a generalized model for various instruments and number of staves can theoretically be produced given a large enough dataset.
 

### 2024-05-09 papers/cs444/paper-cs444.tex
more work!

@@
 However, no paper to date has intended to capture a high accuracy while completely generalizing the space of recognizable music.
+
 Finally, the state-of-the-art for machine-learning-based OMR lies in the implementation of Convolutional Recurrent Neural Networks (CRNN). 
 This model type is preferred over CNN for its ability to learn data as order-dependent sequences- a crucial philosophy in parsing and understanding musical scores in general. 
+However, it should be noted that CNNs are still viable for their non-order-dependent musical problems, such as Nugroho and Zahra's work on individual note and duration recognition ~\cite{nugroho_zahra_2024}.
 
@@
 
-\begin{figure}
-	\begin{subfigure}
-		\centering
-		\includegraphics[width = .8\linewidth]{./figures/lilypond-snippet.png}
-		% \caption{Rendered Lilypond (.ily) file with code}
-		\label{fig:lilypond-snippet}
-	\end{subfigure}
-	~
-	\begin{subfigure}
-		\centering
-		\includegraphics[width = .8\linewidth]{./figures/lilypond-code.png}
-		\label{fig:lilypond-code}
-	\end{subfigure}
-	\caption{Lilypond (.ly) code and rendering, (time signature = 4/4)}
-	\label{fig:lilypond-example}
-\end{figure}
-
 At the music level, rhythmic musical symbols are notated according to their pitches and relative time duration.
@@
 For example, a score with time signature = 2/4 indicates that each measure has two beats, and each beat is the length of a quarter note (hard-coded in the denominator).
-In this case, the line $"c4 d4"$ would represent two quarter (4) notes or a single measure in the provided time signature.
+In this case, the line $"c4 \quad d4"$ would represent two quarter (4) notes or a single measure in the provided time signature.
 Additionally, notes without defined durational values take on the previous note's duration in a line, so a series of equal-duration notes is represented by one durational value.
 Finally, separations between measures are made with $"| \% n"$, marking the end of measure $n$.
-Figure~\ref{fig:lilypond-example} compares the rendering of a single staff of music with its relative Lilypond notation .
+Figure~\ref{fig:lilypond-example} compares the rendering of a single staff of music with its relative Lilypond notation.
+\footnote{The time signature $C$ or "common-time" is another common way of writing 4/4.}
 
-% \begin{lstlisting}
-% trackBchannelB = \relative c {
-% 	r16 c' d e f d e c g'8 c b32 a b16 c8 
-% 	| % 2
-% 	d16 g, a b c a b g d'8 g f32 e f16 g8 
-% 	| % 3
-% }
-% \end{lstlisting}
+\begin{figure}
+	\begin{subfigure}
+		\centering
+		\includegraphics[width = .8\linewidth]{./figures/lilypond-snippet.png}
+	\end{subfigure}
+	~
+	\begin{subfigure}
+		\centering
+		\includegraphics[width = .8\linewidth]{./figures/lilypond-code.png}
+	\end{subfigure}
+	\caption{Lilypond (.ly) code and rendering}
+	\label{fig:lilypond-example}
+\end{figure}
 
@@
 \todo{CRNN model}
-The model in question
+The proposed Image-to-MIDI model is a toy implementation of Shi, et al.'s pioneering work in CRNNs~\cite{shi2015endtoend}.
 
@@
 \todo{dataset specifics}
-Existing MIDI datasets have been narrowed down to strictly two-stave piano works by Johann Sebastian Bach to maintain rhythmic and compositional uniformity, although a generalized model for various instruments and number of staves can theoretically be produced given a large enough dataset.
+MIDI data used for this experiment has been narrowed down to strictly two-stave piano "inventions" by Johann Sebastian Bach ~\cite{bach_midi} to maintain rhythmic and compositional uniformity.
+These works maintain relatively similar quantization schemes (rhythms and durations of notes), are short, and have several clear transcriptions for research, education, and entertainment purposes.
+
+\begin{figure}
+	\begin{subfigure}
+		\centering
+		\includegraphics[width = .8\linewidth]{./figures/invent8.png}
+	\end{subfigure}
+	~
+	\begin{subfigure}
+		\centering
+		\includegraphics[width = .8\linewidth]{./figures/invent4.png}
+	\end{subfigure}
+	\caption{Excerpt of Invention No.4; cropped/prepared rendering of Invention No.8.}
+	\label{fig:bach-example}
+\end{figure}
 

### 2024-05-09 model.py
notebook file

@@ -1,7 +1,7 @@
 import torch
 from torch import nn
 from torch.nn import functional as F
-from torchvision.models import resnet18, ResNet18_Weights
+# from torchvision.models import resnet18, ResNet18_Weights
 from torch import Tensor
 
 
@@ -30,5 +30,24 @@ class CRNN_Network(torch.nn.Module):
         # Transcription
         self.dense = nn.Linear()
 
-    def forward(input):
-        input = 6
+        self.layers = nn.Sequential(
+            self.conv1,
+            self.pool1,
+            self.conv2,
+            self.pool2,
+            self.conv3,
+            self.conv4,
+            self.pool3,
+            self.conv5,
+            self.batch_norm1,
+            self.conv6,
+            self.batch_norm2,
+            self.pool4,
+            self.conv7,
+            self.lstm1,
+            self.lstm2,
+            self.dense
+        )
+
+    def forward(self, input):
+        return self.layers(input)

### 2024-05-10 model.py
cover page

@@ -11,19 +11,19 @@ class CRNN_Network(torch.nn.Module):
 
         self.device = torch.device("cuda" if torch.cuda.is_available() else "cpu")
 
-        self.conv1 = nn.Conv2d()
+        self.conv1 = nn.Conv2d(in_channels=..., out_channels=..., kernel_size=...)
         self.pool1 = nn.MaxPool2d()
-        self.conv2 = nn.Conv2d()
+        self.conv2 = nn.Conv2d(in_channels=..., out_channels=..., kernel_size=...)
         self.pool2 = nn.MaxPool2d()
-        self.conv3 = nn.Conv2d()
-        self.conv4 = nn.Conv2d()
+        self.conv3 = nn.Conv2d(in_channels=..., out_channels=..., kernel_size=...)
+        self.conv4 = nn.Conv2d(in_channels=..., out_channels=..., kernel_size=...)
         self.pool3 = nn.MaxPool2d()
-        self.conv5 = nn.Conv2d()
+        self.conv5 = nn.Conv2d(in_channels=..., out_channels=..., kernel_size=...)
         self.batch_norm1 = nn.BatchNorm2d()
-        self.conv6 = nn.Conv2d()
+        self.conv6 = nn.Conv2d(in_channels=..., out_channels=..., kernel_size=...)
         self.batch_norm2 = nn.BatchNorm2d()
         self.pool4 = nn.MaxPool2d()
-        self.conv7 = nn.Conv2d()
+        self.conv7 = nn.Conv2d(in_channels=..., out_channels=..., kernel_size=...)
         # Map to sequence
         self.lstm1 = nn.LSTM(bidirectional=True)
         self.lstm2 = nn.LSTM(bidirectional=True)

### 2024-05-10 papers/cs444/paper-cs444.tex
cover page
more progress! MOREgit add -A!

@@
 
+\begin{titlepage}
+    \begin{center}
+        \vspace*{1cm}
+
+        \Huge
+        \textbf{Optical Music Recognition for Homophonic LilyPond File Generation}
+
+        \vspace{0.5cm}
+        \LARGE
+
+        \vspace{0.5cm}
+
+        \textbf{Evan Matthews}
+
+        \vspace{0.9cm}
+		\textbf{Executive Summary}
+
+		\vspace{0.5cm}
+		abstract goes here...
+
+		\begin{figure}
+			\centering
+			\includegraphics[width = .8\linewidth]{./figures/cover.png}
+		\end{figure}
+
+        \vfill
+
+        \vspace{0.8cm}
+
+        \Large
+        Department of Computer Science\\
+        University of Illinois Urbana-Champaign\\
+        United States\\
+        May 10th, 2024
+
+    \end{center}
+\end{titlepage}
+
 \title[omr-lilypond-midi]{Optical Music Recognition for Homophonic LilyPond File Generation}
 \author{Evan Matthews}
-\email{evanmm3@illinois.edu}
-%\orcid{1234-5678-9012}
-\affiliation{%
-	\institution{University of Illinois Urbana-Champaign}
-	\country{}
-}
+% \email{evanmm3@illinois.edu}
+% %\orcid{1234-5678-9012}
+% \affiliation{%
+% 	\institution{University of Illinois Urbana-Champaign}
+% 	\country{}
+% }
 
-\begin{abstract}
-\todo{Abstract}
-\end{abstract}
+% \begin{abstract}
+% \todo{Abstract}
+% \end{abstract}
 
@@
 
+For \textbf{RQ1}, I aim to recover noticeable results regarding the CRNN model's ability to generalize sequences of data from scores.
+Hence, whether or not order-dependent training can pick up on note, rhythm or duration sequences will be crucial to my final analysis.
+Next, promising results from \textbf{RQ1} will be compared against existing research to answer \textbf{RQ2}. 
+In particular, the generalized CRNN model's ability to render score data as Lilypond (.ly) files is compared against other notation systems to determine if a more complex Lilypond-based model would be successful.
+Finally, for \textbf{RQ3}, the results of CRNN model training are criticized on what should be generalized or further implemented in order to reliably translate score images into Lilypond data.
+This research question refers to all-encompassing limitations such as in \textbf{1.2} along with experiment-specific limitations.
+
 \section{Background}
@@
 \todo{CRNN model}
-The proposed Image-to-MIDI model is a toy implementation of Shi, et al.'s pioneering work in CRNNs~\cite{shi2015endtoend}.
+The proposed Image-to-MIDI model is a smaller model referencing Shi, et al.'s pioneering work in the creation of CRNNs~\cite{shi2015endtoend}.
+The model itself consists of multiple two-dimensional Convolutional/MaxPooling pairs for the purpose of extracting features from the intial score images.
+Batch normalization layers are also applied to refine data recovered from the feature extraction.
+Lastly, features are pushed through two bidirectional LSTM layers- the recurrent layers of the model- for data prediction with respect to "sequences" of the images.
+In this case, sequences of the image data are the pixel columns of $(595 \times H)$ images, where $H$ is the height of the image dependent on the overall "length" of the music.
+
+\begin{figure}
+	\centering
+	\includegraphics[width = .8\linewidth]{./figures/crnn_arch.png}
+	\caption{General CRNN architecture, as modeled by Shi, et al. ~\cite{shi2015endtoend}}
+	\label{fig:crnn-arch}
+\end{figure}
 
@@
 \section{Results}
+\todo{results}
 
 \section{Discussion}
-\todo{discussion}
+
+The following section details my analysis with respect to previously discussed research questions.
+
+\subsection{Model Generalization (RQ1)}
+
+\subsection{Format Comparison (RQ2)}
+
+\subsection{Model Criticism (RQ3)}
 

### 2024-05-10 model.py
word tokenization

@@ -6,29 +6,29 @@ from torch import Tensor
 
 
 class CRNN_Network(torch.nn.Module):
-    def __init__(self):
+    def __init__(self, in_chans):
         super(CRNN_Network, self).__init__()
 
         self.device = torch.device("cuda" if torch.cuda.is_available() else "cpu")
 
-        self.conv1 = nn.Conv2d(in_channels=..., out_channels=..., kernel_size=...)
-        self.pool1 = nn.MaxPool2d()
-        self.conv2 = nn.Conv2d(in_channels=..., out_channels=..., kernel_size=...)
-        self.pool2 = nn.MaxPool2d()
-        self.conv3 = nn.Conv2d(in_channels=..., out_channels=..., kernel_size=...)
-        self.conv4 = nn.Conv2d(in_channels=..., out_channels=..., kernel_size=...)
-        self.pool3 = nn.MaxPool2d()
-        self.conv5 = nn.Conv2d(in_channels=..., out_channels=..., kernel_size=...)
+        self.conv1 = nn.Conv2d(in_channels=in_chans, out_channels=64, kernel_size=(3,3), stride=1, padding=1, bias=False)
+        self.pool1 = nn.MaxPool2d(kernel_size=(2,2), stride=2)
+        self.conv2 = nn.Conv2d(in_channels=64, out_channels=128, kernel_size=(3,3), stride=1, padding=1, bias=False)
+        self.pool2 = nn.MaxPool2d(kernel_size=(2,2), stride=2)
+        self.conv3 = nn.Conv2d(in_channels=128, out_channels=256, kernel_size=(3,3), stride=1, padding=1, bias=False)
+        self.conv4 = nn.Conv2d(in_channels=256, out_channels=256, kernel_size=(3,3), stride=1, padding=1, bias=False)
+        self.pool3 = nn.MaxPool2d(kernel_size=(1,2), stride=2)
+        self.conv5 = nn.Conv2d(in_channels=256, out_channels=512, kernel_size=(3,3), stride=1, padding=1, bias=False)
         self.batch_norm1 = nn.BatchNorm2d()
-        self.conv6 = nn.Conv2d(in_channels=..., out_channels=..., kernel_size=...)
+        self.conv6 = nn.Conv2d(in_channels=512, out_channels=512, kernel_size=(3,3), stride=1, padding=1, bias=False)
         self.batch_norm2 = nn.BatchNorm2d()
-        self.pool4 = nn.MaxPool2d()
[…]

### 2024-05-10 model.py
getting there...

@@ -19,14 +19,14 @@ class CRNN_Network(torch.nn.Module):
         self.conv4 = nn.Conv2d(in_channels=256, out_channels=256, kernel_size=(3,3), stride=1, padding=1, bias=False)
         self.pool3 = nn.MaxPool2d(kernel_size=(1,2), stride=2)
         self.conv5 = nn.Conv2d(in_channels=256, out_channels=512, kernel_size=(3,3), stride=1, padding=1, bias=False)
-        self.batch_norm1 = nn.BatchNorm2d()
+        self.batch_norm1 = nn.BatchNorm2d(num_features=512)
         self.conv6 = nn.Conv2d(in_channels=512, out_channels=512, kernel_size=(3,3), stride=1, padding=1, bias=False)
-        self.batch_norm2 = nn.BatchNorm2d()
+        self.batch_norm2 = nn.BatchNorm2d(num_features=512)
         self.pool4 = nn.MaxPool2d(kernel_size=(1,2), stride=2)
-        self.conv7 = nn.Conv2d(in_channels=512, out_channels=512, kernel_size=(2,2), stride=1, padding=0, bias=False)
+        self.conv7 = nn.Conv2d(in_channels=512, out_channels=256, kernel_size=(2,2), stride=1, padding=0, bias=False)
         # Map to sequence
-        self.lstm1 = nn.LSTM(hidden_size=256, bidirectional=True)
-        self.lstm2 = nn.LSTM(hidden_size=256, bidirectional=True)
+        self.lstm1 = nn.LSTM(input_size=256, hidden_size=256, bidirectional=True)
+        self.lstm2 = nn.LSTM(input_size=256, hidden_size=256, bidirectional=True)
         # Transcription
         self.dense = nn.Linear(in_features=256, out_features=256)
 

### 2024-05-10 papers/cs444/paper-cs444.tex
.
quantization stuff

@@
         \Huge
-        \textbf{Optical Music Recognition for Homophonic LilyPond File Generation}
+        \textbf{Optical Music Recognition for LilyPond File Generation}
 
@@
 		\vspace{0.5cm}
-		abstract goes here...
+		This report explores the feasibility of Lilypond as a digital music format in the process of optical music recognition and translation for full scores.
+		A simple Convolutional Recurrent Neural Network (CRNN) implementation was trained on Inventions by Johann Sebastian Bach, and test outputs of other Bach works were analyzed for generalization in the Lilypond format.
+		Overall, \todo{results}
 
@@
 
-\title[omr-lilypond-midi]{Optical Music Recognition for Homophonic LilyPond File Generation}
+\title[omr-lilypond-midi]{Optical Music Recognition for LilyPond File Generation}
 \author{Evan Matthews}
@@
 \todo{Introduction citations}
-\cite{contreras2023omrcnn,,andrea2021note}
+\cite{andrea2021note}
 
@@
 
+\begin{itemize}
+	\item \textbf{Symbol Recognition:} a
+	\item \textbf{Music Transcription:} b
+\end{itemize}
+
 \subsection{Lilypond}
@@
 Lastly, features are pushed through two bidirectional LSTM layers- the recurrent layers of the model- for data prediction with respect to "sequences" of the images.
-In this case, sequences of the image data are the pixel columns of $(595 \times H)$ images, where $H$ is the height of the image dependent on the overall "length" of the music.
+In this case, sequences of the image data are the pixel columns of \todo{595?} $(595 \times H)$ images, where $H$ is the height of the image dependent on the overall "length" of the music.
 
@@
 \todo{dataset specifics}
-MIDI data used for this experiment has been narrowed down to strictly two-stave piano "inventions" by Johann Sebastian Bach ~\cite{bach_midi} to maintain rhythmic and compositional uniformity.
+MIDI data used for this experiment has been narrowed down to strictly fifteen two-stave "inventions" by Johann Sebastian Bach ~\cite{bach_midi} to maintain rhythmic and compositional uniformity.
 These works maintain relatively similar quantization schemes (rhythms and durations of notes), are short, and have several clear transcriptions for research, education, and entertainment purposes.
+
+Testing was performed against a handful of arbitrary "preludes" and "fugues," also written by Bach. 
+Limiting the training and testing sets, (especially in a way that limits the ability to calculate an overall accuracy), is purposefully done in the context of this experiment.
+First, there are no expectations for relatively "accurate" recreations of score images in Lilypond. 
+A proper training model would require extensive tokenization for all symbolic details that help to render a Lilypond file, which is outside of the scope of this research endeavor.
+Rather, output is manually analyzed against its correlated image to determine what musical or notational qualities were generalized.
+Second, a broader dataset would drastically complicate the CRNN model's ability to generalize key features of score images.
+Focus on a single composer, musical time period, and music "type" (prelude, waltz, rondo, etc.) allows the model to more quickly generalize aspects of the subset of music, at the cost of potentially overfitting to the subset.
 
@@
 \end{figure}
+
+\subsection{Quantization Rendering}
+Similar to simple transformations performed on images, (rotations, coloring, scaling), output for sheet music determined by technical, but musical variables that don't affect the "performance" of the composition.
+In particular, when score images are rendered from the Bach lilypond files, (trivially translated from the Bach MIDI Index~\cite{bach_midi}), there is a required "quantization" value which affects the rhythms and durations visually.
+This quantization, designated $Q \in \mathbb{Z}$, represents what specific note duration counts as a "beat" within a given score. 
+Figure ~\ref{fig:table-duration} lists the most common duration mappings that occur within this experiment.
+
+\begin{figure}
+	\centering
+	\includegraphics[width = .8\linewidth]{./figures/notevalues.jpeg}
+	\caption{Table of common duration-integer mappings, from Francis Hamzagic ~\cite{hamzagic_2018}}
+	\label{fig:table-duration}
+\end{figure}
+
+Drastic change of $Q$ results in score images that appear muddled or illegible despite conveying the same musical information, as shown in Figure ~\ref{fig:quantization}.
+In order to prevent drastic shifts due to quantization, an optimal quantization $\hat Q$ is calculated with
+
+\begin{figure}
+	\centering
+	\begin{subfigure}
+		\centering
+		\includegraphics[width = .8\linewidth]{./figures/c1.png}
+	\end{subfigure}
+	~
+	\begin{subfigure}
+		\centering
+		\includegraphics[width = .8\linewidth]{./figures/c32.png}
+	\end{subfigure}
+	\caption{Levels of Quantization $[2^0, 2^5]$ in Rendering}
+	\label{fig:quantization}
+\end{figure}
+
+\begin{equation}
+	\hat Q = \max_{t\in \text{MIDI events}} \frac{\tau * 10^{-6}}{t} * n * \text{floor}(d / 4)
+\end{equation}
+
+where $t$ is the number of seconds for a MIDI event to occur, $\tau$ is the provided MIDI tempo in beats per microsecond, and $(n, d)$ are the numerator and denominator, respectively, of the piece's time signature.
+Additionally, possible $\hat Q$ are limited to powers of 2 greater than 0 in order to reduce complication from compound time signatures, (where beats are not represented by durations in Figure ~\ref{fig:table-duration}).
 

### 2024-05-10 papers/cs444/paper-cs444.tex
more work

@@
 \author{Evan Matthews}
-% \email{evanmm3@illinois.edu}
-% %\orcid{1234-5678-9012}
-% \affiliation{%
-% 	\institution{University of Illinois Urbana-Champaign}
-% 	\country{}
-% }
-
-% \begin{abstract}
-% \todo{Abstract}
-% \end{abstract}
 
@@
 \subsection{Initial Proposal}
-\todo{Introduction citations}
-\cite{andrea2021note}
 
@@
 	\item [\textbf{RQ1}] What forms of recognition can be expected from a generalized CRNN implementation?
-	\item [\textbf{RQ2}] How does Lilypond recognition compare against other standard and custom formats?
+	\item [\textbf{RQ2}] How does Lilypond recognition compare against other standard and custom music formats?
 	\item [\textbf{RQ3}] What conflicts currently prevent state-of-the-art models from further generalization?
@@
 	\end{subfigure}
-	\caption{Levels of Quantization $[2^0, 2^5]$ in Rendering}
+	\caption{Levels of Quantization $Q = [1, 32]$ in Rendering}
 	\label{fig:quantization}
@@
 
-The following section details my analysis with respect to previously discussed research questions.
+The following section details my analysis with respect to the previously mentioned research questions (\textbf{RQ1-3}).
 
 \subsection{Model Generalization (RQ1)}
+Given that resulting text output consisted of
 
 \subsection{Format Comparison (RQ2)}
+Lilypond is not, by all means, a perfect music formatting language. 
+Being a format focused on the visual clarity of musical notation, I find that the aforementioned "document level" greatly hinders Lilypond from achieving basic recognition.
+In particular, additional keywords for the representation and position of solely visual elements of a score, 
+along with their necessary language features like brackets and spacers, strongly inhibit basic CRNN models from making noticeable music generalizations.
 
 \subsection{Model Criticism (RQ3)}
+My CRNN implementation, along with the CRNN model for which my research references, are not without barriers preventing consistent musical generalization.
+In staying consistent with my description of the Lilypond language, my critiques against CRNNs are categorized by what should exist in order for music-level and document-level generalizations to occur.
+
+\subsubsection{Music-level Generalization}
+
+\subsubsection{Document-level Generalization}
 
@@
 \todo{conclusions}
+Overall, despite low expectations for workable data, the following experiment and results proved to be quite beneficial in testing the Lilypond format.
 

### 2024-05-10 model.py

@@ -1,7 +1,6 @@
 import torch
 from torch import nn
 from torch.nn import functional as F
-# from torchvision.models import resnet18, ResNet18_Weights
 from torch import Tensor
 
 

### 2024-05-11 papers/cs444/paper-cs444.tex
ahhhh

@@
-\documentclass[review,sigconf]{acmart}
+\documentclass[sigconf]{acmart}
 
@@
 		This report explores the feasibility of Lilypond as a digital music format in the process of optical music recognition and translation for full scores.
-		A simple Convolutional Recurrent Neural Network (CRNN) implementation was trained on Inventions by Johann Sebastian Bach, and test outputs of other Bach works were analyzed for generalization in the Lilypond format.
-		Overall, \todo{results}
+		A simple Convolutional Recurrent Neural Network (CRNN) implementation was trained on Inventions by Johann Sebastian Bach, with intention for test outputs of other Bach works to be analyzed for generalization in the Lilypond format.
+		Overall, an implementation of the Shi, et al. CRNN model was unsuccessful in providing Lilypond predictions for analysis. 
+		However, current state-of-the-art CRNN with other digital music formats provide an extensive survey of what is can be accomplished in full-score optical music recognition.
 
@@
 \subsection{CRNN Model}
-\todo{CRNN model}
 The proposed Image-to-MIDI model is a smaller model referencing Shi, et al.'s pioneering work in the creation of CRNNs~\cite{shi2015endtoend}.
@@
 Lastly, features are pushed through two bidirectional LSTM layers- the recurrent layers of the model- for data prediction with respect to "sequences" of the images.
-In this case, sequences of the image data are the pixel columns of \todo{595?} $(595 \times H)$ images, where $H$ is the height of the image dependent on the overall "length" of the music.
+In this case, sequences of the image data are the pixel columns of $(W \times H)$ images, where $W$ is the pixel width of standard "letter" paper commonly used for scores, and $H$ is the height of the image dependent on the overall "length" of the music.
 
@@
 \subsection{Data Specfications}
-\todo{dataset specifics}
-MIDI data used for this experiment has been narrowed down to strictly fifteen two-stave "inventions" by Johann Sebastian Bach ~\cite{bach_midi} to maintain rhythmic and compositional uniformity.
+MIDI data used for this experiment has been narrowed down to strictly fifteen two-stave "Inventions" by Johann Sebastian Bach ~\cite{bach_midi} to maintain rhythmic and compositional uniformity.
 These works maintain relatively similar quantization schemes (rhythms and durations of notes), are short, and have several clear transcriptions for research, education, and entertainment purposes.
@@
 \section{Results}
-\todo{results}
+Due to technical conflicts with my intended dataset, (variable-height images against Lilypong string data),
+ the CRNN model proved unsuccessful in training and producing any results in the Lilypond formatting language.
+While this result complicates my analysis of Lilypond for machine learning, a meta-analysis of state-of-the-art OMR research with CRNN models still provides significant findings towards the previously mentioned research questions.
 
@@
 
-The following section details my analysis with respect to the previously mentioned research questions (\textbf{RQ1-3}).
+The following section details my meta-analysis of reference work with respect to the previously mentioned research questions (\textbf{RQ1-3}).
 
 \subsection{Model Generalization (RQ1)}
-Given that resulting text output consisted of
+Given the visual encoding aspects of Lilypond, a simple CRNN model would struggle to accomplish several layers of recognition at once.
+However, multiple simple CRNNs would be successful in learning the Lilypond file schematic and musical information provided that both levels are tokenized and separated.
+The work of Nugroho and Zahra, for instance, validates the ability of a simple CNN architecture to solely recognize musical symbols in context ~\cite{nugroho_zahra_2024}.
+
+With additional features and data preparation, CRNN models are shown to accurately parse and transcribe full scores of monophonic\footnote{\textit{monophonic} refers to having one "voice" of musical information.} and homophonic~\footnote{\textit{homophonic} refers to having multiple voices of music sharing the same note rhythms and durations.} music ~\cite{contreras2023omrcnn}.
+In turn, the additional preparation, which forcibly standardizes the data to assist in training, sterilizes the diverse nuances that come with all possible sheet music, 
+including physical aging, printing error, different notation fonts, sizes and shapes of scores, and the immense dictionary of musical symbols, (standardized or unique to specific pieces of music).
 
@@
 
+More complex CRNN models and state-of-the-art work currently provides towards support for Lilypond in this context.
+Mayer, et al.'s \textit{Linearized MusicXML}~\cite{mayer2024practical}, a specially-designed superset of MusicXML, demonstrates how complex CRNNs can accurately generalize document-level and music-level data.
+
 \subsection{Model Criticism (RQ3)}
-My CRNN implementation, along with the CRNN model for which my research references, are not without barriers preventing consistent musical generalization.
+My CRNN implementation, along with the CRNN model for which my research references, are not without active barriers preventing consistent musical generalization.
 In staying consistent with my description of the Lilypond language, my critiques against CRNNs are categorized by what should exist in order for music-level and document-level generalizations to occur.

### 2024-05-11 papers/cs444/paper-cs444.tex
paper is done

@@
-\documentclass[sigconf]{acmart}
+\documentclass[nonacm, sigconf]{acmart}
 
@@
 \subsection{Optical Music Recognition}
-\todo{OMR background}
+Current research in Optical Music Recognition (OMR) can be put in one of two categories:
 
 \begin{itemize}
-	\item \textbf{Symbol Recognition:} a
-	\item \textbf{Music Transcription:} b
+	\item \textbf{Symbol Recognition:} the ability for a model to recognize individual symbols on a score.
+	\item \textbf{Music Transcription:} the ability for a model to recognize and transcribe multiple symbols or long passages into a new musical format.
 \end{itemize}
+
+The problem of Symbol Recognition is nearly identical to that of text recognition with the exception of a separate set of possible symbols.
+Additionally, unlike situations where text is perfectly aligned for recognition, musical symbols need to be recognized with respect to their shape and position in order to achieve a high accuracy.
+All notes, for example, have unique pitches dependent on their placement in a score, (on a "line" or "space" in a staff, above or below a staff).
+The playback of notes is also determined by surrounding symbols such as accidentals (natural, sharp, flat), rhythmic symbols (the type of duration, a "dot" for an additional half of that note's duration).
+Finally, the highest level of context is staff-related information, which includes clefs (treble, bass, alto, tenor; determining the relation between note placement and audible pitches), time signatures and key signatures (for adding context for an entire piece of music).
+
+On the other hand, Music Transcription entirely of the Symbol Recognition problem, (iteratively for some amount of sheet music) as well as problems with correlating recognized symbols in sequence.
+On its own, Symbol Recognition could recognize every component of a piece of music, but these results lack the ordered context necessary for transcription.
+State-of-the-art transcription through machine learning, as a result, has been dominated by the Convolutional Recurrent Neural Network (CRNN). 
+Unlike CNNs, CRNNs make use of recurrent layers, (such as bidirectional LSTM layers in Shi, et al. ~\cite{shi2015endtoend}) in order to contextualize sequences of data forwards and backwards.
+These layers, along with attention cells focused on deriving pitch, rhythm and durational context, allow for CRNNs to recognize and output ordered sequences of logical musical data.
+
+Additionally, it is important to recognize that musical data sequences are ordered along multiple dimensions.
+Contreras, et al. provides an excellent diagram of this ordering in their work with respect to how notes have a separate "ordering" from other symbols.
+State-of-the-art CRNNs, in turn, are able to recognize this multi-dimensional ordering to a point, (see 5.3.1) so long as specific elements of the music are kept constant.
+In the case of most OMR research, the rhythms and durations are kept constant across multiple voices, (unless only a single voice is used).
+
+\begin{figure}
+	\centering
+	\includegraphics[width = .8\linewidth]{./figures/ordering.png}
+	\caption{multi-dimensional ordering in sheet music ~\cite{contreras2023omrcnn}.}
+	\label{fig:ordering}
+\end{figure}
 
@@
 
-Testing was performed against a handful of arbitrary "preludes" and "fugues," also written by Bach. 
+Testing is performed against a handful of arbitrary "preludes" and "fugues," also written by Bach. 
 Limiting the training and testing sets, (especially in a way that limits the ability to calculate an overall accuracy), is purposefully done in the context of this experiment.
@@
 
-With additional features and data preparation, CRNN models are shown to accurately parse and transcribe full scores of monophonic\footnote{\textit{monophonic} refers to having one "voice" of musical information.} and homophonic~\footnote{\textit{homophonic} refers to having multiple voices of music sharing the same note rhythms and durations.} music ~\cite{contreras2023omrcnn}.
+With additional features and data preparation, CRNN models are shown to accurately parse and transcribe full scores of monophonic\footnote{\textit{monophonic} refers to having one voice of musical information.} and homophonic~\footnote{\textit{homophonic} refers to having multiple voices of music sharing the same note rhythms and durations.} music ~\cite{contreras2023omrcnn}.
 In turn, the additional preparation, which forcibly standardizes the data to assist in training, sterilizes the diverse nuances that come with all possible sheet music, 
@@
 My CRNN implementation, along with the CRNN model for which my research references, are not without active barriers preventing consistent musical generalization.
-In staying consistent with my description of the Lilypond language, my critiques against CRNNs are categorized by what should exist in order for music-level and document-level generalizations to occur.
+In staying consistent with my description of the Lilypond language, my critiques against CRNNs are categorized by what further work should be done in order for music-level and document-level generalizations to occur.
 
 \subsubsection{Music-level Generalization}
+The primary issue holding current CRNN models back is their inability to generalize to polyphonic~\footnote{\textit{polyphonic} refers to having multiple voices of music with different rhythms and durations.} music.
+That is, state-of-the-art models fail to solve the problem of recognizing and transcribing several different "voices" of music at the same time.
+Solutions related to the separation of these voices, (i.e. removing sequences as they are recognized from the original score), approach the homophonic algorithm, but they also introduce problems.
+For instance, when a voice is removed, determining whether intersecting voices were affected is rather difficult.
+
+Additionally, a majority of OMR research relies on datasets consisting of digital or digitized musical scores.
+This choice for data is fair when considering that OMR models require upwards of hundreds of scores for proper training~\cite{mayer2024practical}, but it also limits the diversity of scores and contexts that ML transcription can be used for.
+In particular, standards for scores have changed several times over centuries of music history, yet the formatting of historical scores is often left behind in modern transcriptions.
+The solution to this issue, while partly related to current models, lies in the amount of trainable data across all eras of classical music and the ability to notate historical scores.
+Modern notation software, while addressing current standards, thankfully includes typing for historical symbols, but these symbols have yet to be properly implemented in popular formatting languages.
+Historical datasets are also quite possible, albeit as scanned images with no manually-transcribed digital reference as of the current day.
 
 \subsubsection{Document-level Generalization}
+At the document level, the greatest issue plaguing OMR research is the lack of a formatting language standard, or lack of similarity between all formatting languages.
+Every formatting language has separate means of conveying equivalent musical information based on the user's needs, (i.e.
+MIDI for real-time data, MusicXML for web-related music, Lilypond for polished layouts).
+However, as the needs of users vary in the numerous niches of music production, technology and composition, digital music encoding and datasets slowly slow the possibility of standardization.
+While MIDI and MusicXML are currently the most popular for datasets, both lack consistency in rendering sheet music across all digital notation platforms.
+I believe consistency on this front is crucial for reducing variability in musical machine learning datasets.
+In particular, sheet music should "look the same" regardless of the formatting language it is imported from.
 
 \section{Conclusions}
-\todo{conclusions}
-Overall, despite low expectations for workable data, the following experiment and results proved to be quite beneficial in testing the Lilypond format.
+Overall, despite unsuccessful training of a simple CRNN model on Lilypond data, the following meta-analysis proved to be quite beneficial in analyzing the Lilypond formatting language.
+Existing research from several CRNN papers confirms the viability of Lilypond as a format for machine learning, and I believe a combination of this, tokenization for layout-related encoding, 
+and active development of the language would allow it to thrive in current research. 
+I'm especially interested in seeing Lilypond extend into OMR research as its backend contains useful tools for conversion to MIDI and MusicXML, formats with plenty of existing datasets but higher complexity in generalization.
+
+In the future, I would like to focus on a tokenization system for future Lilypond ML models, as this sudden obstacle was the reason for the experiment being unsuccessful.
+Additionally, given my interest in music notation and conversion to related formats, I am hoping to generate non-text-related outputs from digital music data, including audio for performances,
+video for artistic interpretation, and more scientific metrics to bridge the gap between "musical emotion" and psychology.
+
+\section{Conflict of Interest}
+
+My ongoing plans for thesis work, advised by Paris Smaragdis, revolve around the generation of audio conditioned on score images. 
+While this report actively assists in my understanding of the correlations between digital music formats and computer vision, I do not believe that it trivializes any future work in this topic.
 

### 2024-05-11 papers/cs444/paper-cs444.tex
paper double-checked

@@
 Machine learning models, in turn, have been a crucial step towards reducing these inconsistencies. 
-Their ability to learn the nonlinearities and artistic qualities that otherwise plague audio computation have allowed for noteworthy advancements such as the WaveNet~\cite{oord2016wavenet}. 
+Their ability to learn the nonlinearities and artistic qualities that otherwise plague audio computation have allowed for noteworthy advancements such as WaveNet~\cite{oord2016wavenet}. 
 The remaining issue in the process of sound generation is the amount of data available to train with. 
@@
 
-Given the vast complexity of western music notation and intricacies in notation/time alignment, the proposed model has a pessimistic approach. 
+Given the vast complexity of classical western music notation and intricacies in notation/time alignment, the proposed model maintins a rather pessimistic approach. 
 That is, given a small subset of musical data (\textit{among all notated music}) and some model, results can be expected to only match expected outputs through one or more of: pitch, rhythm, duration, and overall formatting.
@@
 	\item Software-specific formats from \textit{Musescore}, \textit{Sibelius} and \textit{Finale},
-	\item \textit{KERN} from the Humdrum tool-set ~\cite{contreras2023omrpian},
+	\item \textit{KERN} from the Humdrum tool-set ~\cite{contreras2023omrpiano},
 	\item Mayer, et al.'s \textit{Linearized MusicXML} ~\cite{mayer2024practical}, and
@@
 
-Despite low research expectations, I believe that a few questions can be posed and answered for the purpose of bounding requirements on a larger, all-encompassing Image-to-MIDI model:
+Despite low expectations for training and results, I believe that a few questions can be posed and answered for the purpose of bounding requirements on a larger, all-encompassing Image-to-MIDI model:
 
@@
 \subsection{Optical Music Recognition}
-Current research in Optical Music Recognition (OMR) can be put in one of two categories:
+Current research in Optical Music Recognition (OMR) can be placed in one of two categories:
 
@@
 
-On the other hand, Music Transcription entirely of the Symbol Recognition problem, (iteratively for some amount of sheet music) as well as problems with correlating recognized symbols in sequence.
+On the other hand, Music Transcription encompasses the entire Symbol Recognition problem, (iteratively for some amount of sheet music) as well as problems with correlating recognized symbols in sequence.
 On its own, Symbol Recognition could recognize every component of a piece of music, but these results lack the ordered context necessary for transcription.
@@
 Pitches are also proceeded by a rhythmic value representing its duration in time.
-These values are typically powers of two, (but can technically be any positive floating-point value), and they dictate how long a pitch is played with respect to the piece's time signature.
+These values are typically powers of two, (but can technically be any positive integer), and they dictate how long a pitch is played with respect to the piece's time signature.
 For example, a score with time signature = 2/4 indicates that each measure has two beats, and each beat is the length of a quarter note (hard-coded in the denominator).
@@
 	\end{subfigure}
-	\caption{Lilypond (.ly) code and rendering}
+	\caption{Lilypond (.ly) code and rendering.}
 	\label{fig:lilypond-example}
@@
 	\includegraphics[width = .8\linewidth]{./figures/crnn_arch.png}
-	\caption{General CRNN architecture, as modeled by Shi, et al. ~\cite{shi2015endtoend}}
+	\caption{General CRNN architecture, as modeled by Shi, et al. ~\cite{shi2015endtoend}.}
 	\label{fig:crnn-arch}
@@
 \subsection{Quantization Rendering}
-Similar to simple transformations performed on images, (rotations, coloring, scaling), output for sheet music determined by technical, but musical variables that don't affect the "performance" of the composition.
+Similar to simple transformations performed on images, (rotations, coloring, scaling), output for sheet music is determined by both technical and musical constraints that don't affect the "performance" of the composition.
 In particular, when score images are rendered from the Bach lilypond files, (trivially translated from the Bach MIDI Index~\cite{bach_midi}), there is a required "quantization" value which affects the rhythms and durations visually.
-This quantization, designated $Q \in \mathbb{Z}$, represents what specific note duration counts as a "beat" within a given score. 
+This quantization, designated $Q \in \mathbb{Z}^+$, represents what specific note duration counts as a "beat" within a given score. 
 Figure ~\ref{fig:table-duration} lists the most common duration mappings that occur within this experiment.
@@
 	\includegraphics[width = .8\linewidth]{./figures/notevalues.jpeg}
-	\caption{Table of common duration-integer mappings, from Francis Hamzagic ~\cite{hamzagic_2018}}
+	\caption{Table of common duration-integer mappings, from Francis Hamzagic ~\cite{hamzagic_2018}.}
 	\label{fig:table-duration}
@@
 	\end{subfigure}
-	\caption{Levels of Quantization $Q = [1, 32]$ in Rendering}
+	\caption{Levels of quantization $Q = (1, 32)$ in rendering, respectively.}
 	\label{fig:quantization}
@@
 \subsection{Model Criticism (RQ3)}
-My CRNN implementation, along with the CRNN model for which my research references, are not without active barriers preventing consistent musical generalization.
-In staying consistent with my description of the Lilypond language, my critiques against CRNNs are categorized by what further work should be done in order for music-level and document-level generalizations to occur.
+My CRNN implementation, along with the CRNN models for which my research references, are not without active barriers preventing consistent musical generalization.
+In staying consistent with my description of the Lilypond formatting language, my critiques against CRNNs are categorized by what further work should be done in order for music-level and document-level generalizations to occur.
 
@@
 MIDI for real-time data, MusicXML for web-related music, Lilypond for polished layouts).
-However, as the needs of users vary in the numerous niches of music production, technology and composition, digital music encoding and datasets slowly slow the possibility of standardization.
-While MIDI and MusicXML are currently the most popular for datasets, both lack consistency in rendering sheet music across all digital notation platforms.
+However, as the needs of users vary in the numerous niches of music production, technology and composition, digital music encoding and datasets slowly diminish the possibility of standardization.
+While MIDI and MusicXML are currently the most popular for datasets, both formats lack consistency in rendering sheet music across all digital notation platforms.
 I believe consistency on this front is crucial for reducing variability in musical machine learning datasets.
